---
title: "Full Simulation Example with CDMConnector"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full Simulation Example with CDMConnector}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

OncoRegimenTester generates synthetic oncology drug exposure data for testing
regimen detection algorithms. This vignette walks through a full simulation
and shows how to work with the output alongside a real OMOP CDM dataset
(the Eunomia GiBleed example from CDMConnector).

The idea: in real-world data you only see drug exposures. You never directly
observe which regimen a patient is on. Regimen detection algorithms try to
reconstruct that from the drug-level data. This simulator lets you test those
algorithms against known ground truth.

## Setup

```{r setup}
library(OncoRegimenTester)
library(data.table)
```

## Step 1: Load and validate parameters

The simulator is driven by CSV parameter files. The included mCRC (metastatic
colorectal cancer) example defines regimens like FOLFOX, FOLFIRI, CAPOX, and
pembrolizumab monotherapy.

```{r load-params}
paramDir <- system.file("example_params", "mCRC", package = "OncoRegimenTester")
pack <- loadParameterPack(paramDir)

# Check what we loaded
validation <- validateParameterPack(pack)
cat(paste(validation$summary, collapse = "\n"))
```

```{r show-warnings}
if (length(validation$warnings) > 0) {
  cat("\nWarnings:\n")
  cat(paste(validation$warnings, collapse = "\n"))
} else {
  cat("\nNo validation warnings.")
}
```

## Step 2: Inspect the regimen definitions

Before generating, let's see what regimens we're working with:

```{r regimen-catalog}
pack$regimenCatalog[, .(regimen_id, regimen_name, regimen_type, is_maintenance)]
```

And which drugs belong to each regimen:

```{r regimen-drugs}
pack$regimenDrugMap[, .(regimen_id, drug_feature_id, drug_role, required_flag)]
```

## Step 3: Generate synthetic data

Now generate 200 patients with some realistic noise:

```{r generate}
config <- simulatorConfig(
  paramDir = paramDir,
  outputDir = tempfile("oncoregimen_"),
  nPatients = 200L,
  seed = 42L,
  noise = noiseConfig(
    probabilityJitterSd = 0.1,
    durationJitterDays = 3L,
    cycleJitterDays = 1L,
    missingnessRate = 0.03
  )
)

result <- runSimulation(config)
```

## Step 4: Examine the generated data

### Patients

```{r patients}
head(result$patients)

# Profile distribution
result$patients[, .N, by = profile_id]
```

### Drug exposures

The drug exposures are the primary output -- this is what a regimen detection
algorithm would see:

```{r drug-exposures}
de <- result$therapy$drugExposures
cat(sprintf("Total drug exposures: %d\n", nrow(de)))
cat(sprintf("Unique drugs: %d\n", length(unique(de$drug_feature_id))))

# Exposures per drug
de[, .N, by = drug_feature_id][order(-N)]
```

### Ground truth vs detected regimens

The ground truth is what the simulator intended:

```{r ground-truth}
result$therapy$groundTruthEpisodes[1:10, ]
```

The detected regimens are what the reference detector recovered from the
drug exposures alone:

```{r detected}
result$detected$episodes[1:10, ]
```

### Compare ground truth to detected

```{r comparison}
gtEpisodes <- nrow(result$therapy$groundTruthEpisodes)
detEpisodes <- nrow(result$detected$episodes)

cat(sprintf("Ground truth episodes: %d\n", gtEpisodes))
cat(sprintf("Detected episodes: %d\n", detEpisodes))

# How often does the detector correctly identify the regimen?
gt <- result$therapy$groundTruthEpisodes
det <- result$detected$episodes

# Merge on patient_id and overlapping time windows
gt[, gt_key := paste(patient_id, regimen_id, sep = "_")]
det[, det_key := paste(patient_id, detected_regimen_id, sep = "_")]

matchRate <- length(intersect(gt$gt_key, det$det_key)) / nrow(gt)
cat(sprintf("Approximate regimen match rate: %.1f%%\n", matchRate * 100))
```

## Step 5: Working with CDMConnector (Eunomia GiBleed)

If you have CDMConnector installed, you can compare the structure of the
synthetic data with a real OMOP CDM dataset.

```{r eunomia-setup, eval = requireNamespace("CDMConnector", quietly = TRUE) && requireNamespace("duckdb", quietly = TRUE)}
library(CDMConnector)
library(DBI)
library(duckdb)

# Load the Eunomia GiBleed dataset
con <- dbConnect(duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")

# Look at real drug exposures for comparison
realDe <- as.data.table(head(cdm$drug_exposure, 1000))
cat(sprintf("Eunomia drug_exposure columns: %s\n",
            paste(names(realDe), collapse = ", ")))
cat(sprintf("Eunomia drug_exposure rows (sample): %d\n", nrow(realDe)))
```

```{r eunomia-compare, eval = requireNamespace("CDMConnector", quietly = TRUE) && requireNamespace("duckdb", quietly = TRUE)}
# The synthetic data has the same structural columns
syntheticDe <- result$therapy$drugExposures
cat(sprintf("\nSynthetic drug_exposure columns: %s\n",
            paste(names(syntheticDe), collapse = ", ")))

# Key difference: synthetic data uses drug_feature_id (grouped concept)
# instead of individual drug_concept_id values. This is by design --
# the simulator works at the feature level for interpretability.
```

### Write synthetic data to a CDM

You can write OncoRegimenTester output directly into a CDMConnector CDM:

```{r write-to-cdm, eval = requireNamespace("CDMConnector", quietly = TRUE) && requireNamespace("duckdb", quietly = TRUE)}
# Create a fresh in-memory database for synthetic data
synCon <- dbConnect(duckdb())

# Write the OMOP tables
omopTables <- renderOmop(
  result$patients, result$therapy, result$supporting,
  result$detected, result$config
)

# Write person and drug_exposure to the database
dbWriteTable(synCon, "person", as.data.frame(omopTables$person))
dbWriteTable(synCon, "drug_exposure", as.data.frame(omopTables$drug_exposure))
dbWriteTable(synCon, "observation_period", as.data.frame(omopTables$observation_period))

cat("Synthetic OMOP tables written to DuckDB.\n")

# Query it like a normal CDM
syntheticPersonCount <- dbGetQuery(synCon, "SELECT COUNT(*) AS n FROM person")
cat(sprintf("Synthetic person count: %d\n", syntheticPersonCount$n))

syntheticDeCount <- dbGetQuery(synCon, "SELECT COUNT(*) AS n FROM drug_exposure")
cat(sprintf("Synthetic drug_exposure count: %d\n", syntheticDeCount$n))

dbDisconnect(synCon, shutdown = TRUE)
```

```{r cleanup, eval = requireNamespace("CDMConnector", quietly = TRUE) && requireNamespace("duckdb", quietly = TRUE)}
dbDisconnect(con, shutdown = TRUE)
```

## Step 6: Validation report

The validation report compares observed frequencies to parameter targets:

```{r validation}
cat(paste(result$validation$report, collapse = "\n"))
```

## Summary

This vignette demonstrated:

1. **Loading parameter CSVs** that define regimen templates, drug schedules,
   and patient profiles
2. **Generating synthetic drug exposures** with realistic timing, jitter,
   and missingness
3. **Comparing ground truth to detected regimens** -- the core evaluation loop
4. **Integrating with CDMConnector** to work alongside real OMOP CDM datasets
5. **Validating** that generated data matches parameter targets

The simulator's purpose is to create test data for regimen detection. The drug
exposures are intentionally messy because real data is messy. A good regimen
detection algorithm should recover the ground truth regimens despite the noise.
